Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-20 manu-draft.Rmd
working on methods revision, mult comp tests for sav, mcr to supp

load(file = here('tables/wqsumtab.RData'))
load(file = here('tables/stktab.RData'))
load(file = here('tables/wqcmptab.RData'))
load(file = here('tables/savfotab.RData'))
load(file = here('tables/mcrfotab.RData'))

@@

## Seagrass and macroalgae trends

@@

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 1), 
    mo = month(date)
  )
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1sg <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = mean(foest)) %>% 
  deframe

fomoarea1apr <- fomoarea1 %>% 
  filter(mo %in% 4) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea1jul <- fomoarea1 %>% 
  filter(mo %in% 7) %>% 
  select(taxa,foest) %>% 
  deframe
fomoarea3apr <- fomoarea3 %>% 
  filter(mo %in% 4) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea3jul <- fomoarea3 %>% 
  filter(mo %in% 7) %>% 
  select(taxa,foest) %>% 
  deframe
fomoarea1jun <- fomoarea1 %>% 
  filter(mo %in% 6) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea3jun <- fomoarea3 %>% 
  filter(mo %in% 6) %>% 
  select(taxa, foest) %>% 
  deframe

# kw pvals
cyanoarea3 <- mcrfotab %>% 
  mutate(
    Area = ifelse(Area == '', NA, Area)
  ) %>% 
  fill(Area) %>% 
  filter(Area == 3 & `Macroalgae group` == 'Cyanobacteria') %>% 
  pull(pval) %>% 
  as.numeric %>% 
  round(3)
```

A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through September, each visited on average `r avevis` times per month. Turtle grass (*T. testudinum*) was the dominant seagrass species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].  At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.  Macroalgae observed along the transects also varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.  Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.

Temporal progression of seagrasses and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).  Transect S3T6 is located less than one kilometer to the north of Port Manatee.  The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.  Overall abundance of seagrass did not change dramatically from April 22nd through September.  However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.  *Dapis sp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa at the end of September.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of seagrass and macroalgae trends across all transects.  No transects were sampled in area 2 to the north of Piney Point and no transects were sampled past September in Area 1.  Area 1 had generally stable frequency occurrence over time of the three primary seagrass species observed in the bay, with turtle grass (*T. testudinum*) occurring slightly more often (average frequency occurrence across months `r fomoarea1sg['Thalassia testudinum']`\%) than shoal grass (*H. wrightii*, `r fomoarea1sg['Halodule wrightii']`\%)) or manatee grass (*S. filiforme*, `r fomoarea1sg['Syringodium filiforme']`\%)).

, although a slight decrease over time in shoal grass (*H. wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea1jul['Halodule wrightii']`\%).  A similar decline in coverage of shoal grass was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%). Frequency occurrence of manatee grass (*S. syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, July frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).

Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area. Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3). Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Green macroalgae had the second lowest frequency occurrence, followed by Brown macroalgae that was only observed at one transect.

Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrfotab) and \@ref(tab:savfotab) further suggested that seagrass trends were minor compared to those of macroalgae.  Results by month showed that seagrass frequency occurrence, by species and area, had insignificant changes ($p > 0.05$) over the study period.  Although trends were not significant, estimated medians in each month in Table \@ref(tab:savfotab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.  These changes were more pronounced in area 3 as compared to area 1.  For macroalgae (Table \@ref(tab:mcrfotab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).  Coverage of cyanobacteria in area 1 during June was significantly higher than coverage in April.  No significant differences between months were observed for the green macroalgae group.

@@

```{r mcrfotab}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savfotab}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

>>>>

load(file = here('tables/wqsumtab.RData'))
load(file = here('tables/stktab.RData'))
load(file = here('tables/wqcmptab.RData'))

@@

## Macroalgae and seagrass trends

@@

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 1), 
    mo = month(date)
  )
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1ave <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 1)) %>% 
  deframe
fomoarea3ave <- fomoarea3 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 1)) %>% 
  deframe

fomoarea1apr <- fomoarea1 %>% 
  filter(mo %in% 4) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea1jul <- fomoarea1 %>% 
  filter(mo %in% 7) %>% 
  select(taxa,foest) %>% 
  deframe
fomoarea3apr <- fomoarea3 %>% 
  filter(mo %in% 4) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea3jul <- fomoarea3 %>% 
  filter(mo %in% 7) %>% 
  select(taxa,foest) %>% 
  deframe
fomoarea1jun <- fomoarea1 %>% 
  filter(mo %in% 6) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea3jun <- fomoarea3 %>% 
  filter(mo %in% 6) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea1sep <- fomoarea1 %>% 
  filter(mo %in% 9) %>% 
  select(taxa, foest) %>% 
  deframe
fomoarea3oct <- fomoarea3 %>% 
  filter(mo %in% 10) %>% 
  select(taxa, foest) %>% 
  deframe
```

A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.  Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.  Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.  For seagrasses, turtle grass (*T. testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].  At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.

Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.  *Dapis sp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change dramatically from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.  No transects were sampled in area 2 to the north of Piney Point and no transects were sampled past September in Area 1.  Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).  Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3). Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in area 1, `r fomoarea3oct['Green']`\% in October in area 3).  Brown macroalgae was only observed at one transect. For seagrass, both areas had generally stable total frequency occurrence. Turtle grass (*T. testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall area 3), compared to shoal grass (*H. wrightii*, `r fomoarea1ave['Halodule wrightii']`\% area 1, `r fomoarea3ave['Halodule wrightii']`\% area 3) and manatee grass (*S. filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% area 1, `r fomoarea3ave['Syringodium filiforme']`\% area 3).  Slight changes in frequency occurrence in area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass. Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).

@@

### 2021-10-20 supplement.Rmd
working on methods revision, mult comp tests for sav, mcr to supp

load(file = here('tables/mcrabutab.RData'))
load(file = here('tables/savabutab.RData'))
```

@@

>>>>

load(file = here('tables/mcrfotab.RData'))
load(file = here('tables/savfotab.RData'))
load(file = here('tables/mcrabutab.RData'))
load(file = here('tables/savabutab.RData'))
```

@@

```{r mcrfotab}
cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.  Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- mcrfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```

```{r savfotab}
cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.  Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.  Multiple comparisons with Mann-Whitney U tests (Comp. column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.  Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs. Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79. ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
totab <- savfotab %>% 
  select(-pval)
knitr::kable(totab, booktabs = T, caption = cap.val)
```
